---
title: "MCMC Method Comparisons: Beta-Gen-t-EGARCH"
output: 
  rmarkdown::html_vignette:
vignette: >
  %\VignetteIndexEntry{MCMC Method Comparisons: Beta-Gen-t-EGARCH}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{AdMit, coda, moments}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.show = 'hold',
  fig.width = 7,
  fig.height = 5,
  fig.align = 'center'
)
```

# MCMC Method Comparisons: Beta-Gen-t-EGARCH

This document contains the R code to reproduce the plots and statistical analysis presented in section 4.1 of the thesis "Bayesian Inference for Generalized Autoregressive Score Models".

```{r lib}
library(BayesianGAS)
```

```{r seed_and_constants}
set.seed(100)
kFiveYrsIdx <- 3769
kNumParams  <- 6
```

## Load data 

```{r load_data}
data("SP500", package = "BayesianGAS")
dates   <- as.Date(rev(spData$Date)[-(1:kFiveYrsIdx)])
returns <- ts(
  diff(rev(spData$Adj.Close))[-(1:(kFiveYrsIdx - 1))],
  start = c(2012, 4, 16),
  frequency = 254
)
```

### Plot Returns
```{r}
leg <- "S&P 500 Returns"
par(mfrow = c(1, 1))
plot(dates, returns, type = "l", xaxt = "n", xlab = "Years", ylab = "Return")
legend("topleft", legend = leg, bty = "n", lty = 1)
axis.Date(1, at = seq(min(dates), max(dates), by = "4 mon"), format = "%Y-%m")
```

### Set data attributes
```{r set_data_attr}
exKurt <- moments::kurtosis(returns) - 3
# Based on student-t second and fourth order moments
ltScale <- log(stats::sd(returns) * sqrt((6 / exKurt + 2) / (6 / exKurt + 4)))
numObs <- length(returns)
```

## Maximum Likelihood (ML) estimation 

```{r ML_estimation_betategarch}
initParams  <- c(
  omega = ltScale,
  A = 0.05,
  B = 0.95,
  mu = mean(returns),
  etaBar = 0.2,
  upsilon = 2
)
startTime <- Sys.time()
betaGen <- FitML(
  model = new(BetaGenTEGARCH),
  initParams = initParams,
  y = returns,
  f1 = ltScale,
  method = 'BFGS',
  control = list(maxit = 1e5),
  hessian = TRUE,
  verbose = TRUE
)
endTime <- Sys.time()
timeML <- difftime(endTime, startTime, units = 'secs')
cat("ML Time: ", timeML, sprintf(" seconds\n"))
```

### Plot volatility
```{r}
betaGen$SetParams(betaGen$ParamsML)
volsGen <- betaGen$VolFilter(returns, ltScale)
par(mfrow = c(1, 1))
plot.ts(cbind(abs(returns - betaGen$Mu), volsGen), col = c("black", "red"),
        lwd = c(1, 1.5), plot.type = "single", ylab = "Volatility")
leg <- c(bquote(GAS - t ~ (logL = .(round(betaGen$LogLValML, 4)))))
legend("topleft", legend = leg, lty = c(1), lwd = c(1.5), col = c("red"),
       bty = "n")

```


## Markov Chain Monte Carlo (MCMC)

```{r}
iter <- 4e4
priorStack <- new(
  PriorStack,
  rep("ImproperUniform", kNumParams),
  list(
    c(-Inf, Inf),
    c(-Inf, Inf),
    c(-1, 1),
    c(-Inf, Inf),
    c(0, 0.5),
    c(0, Inf)
  )
)
```

### Random Walk Metropolis Hastings (RWMH)
```{r}
startTime <- Sys.time()
firstWarmUpRWMH <- RWMH(
  "BetaGenTEGARCH",
  priorStack,
  y = as.matrix(returns),
  f1 = ltScale,
  initParams = initParams,
  sigma = diag(kNumParams),
  iter = 100,
  stepsize = 0.01
)
secondWarmUpRWMH <- RWMH(
  "BetaGenTEGARCH",
  priorStack,
  y = as.matrix(returns),
  f1 = ltScale,
  initParams = initParams,
  sigma = cov(firstWarmUpRWMH),
  iter = 1000,
  stepsize = 1
)
drawsRWMH <- RWMH(
  "BetaGenTEGARCH",
  priorStack,
  y = as.matrix(returns),
  f1 = ltScale,
  initParams = initParams,
  sigma = cov(secondWarmUpRWMH),
  iter = iter,
  stepsize = 1
)
endTime <- Sys.time()
timeRWMH <- difftime(endTime, startTime, units = 'secs')
cat("RWMH Time: ", timeRWMH, sprintf(" seconds\n"))
```

### Adaptive Mixture of Student-t Distributions (AdMit)-MH
```{r}
logOffset <- 250 - betaGen$LogLValML # constant needed to prevent under/overflow
startTime <- Sys.time()
tries <- 0
# AdMit Optimization occsianlly fails,
# give it a few tries as randomness might be on your side.
while (tries < 5) {
  possibleError <- tryCatch(
    outAdMit <- AdMit::AdMit(
      VectorizedPosterior,
      mu0 = initParams,
      control = list(df = 1, trace = TRUE),
      modelStr = "BetaGenTEGARCH",
      priorStack = priorStack,
      y = as.matrix(returns),
      f1 = ltScale,
      logOffset = logOffset
    ),
    error = function(e) e
  )

  if (!inherits(possibleError, "error")) {
    cat("AdMit tries: ", tries + 1, sprintf("\n"))
    break
  }else{
    tries <- tries + 1
    cat("ERROR: ", sprintf("\n"))
    print(possibleError)
  }
}
if (tries == 5) {
  sprintf("AdMit optimization failed 5 times. Try different settings.")
}

betaGenAdMitMH <- AdMit::AdMitMH(
  N = iter,
  KERNEL = VectorizedPosterior,
  modelStr = "BetaGenTEGARCH",
  priorStack = priorStack,
  y = as.matrix(returns),
  f1 = ltScale,
  logOffset = logOffset,
  mit = outAdMit$mit
)
endTime <- Sys.time()
timeAdMit <- difftime(endTime, startTime, units = 'secs')
cat("AdMit-MH Time: ", timeAdMit, sprintf(" seconds\n"))
cat("AdMit-MH Accept Rate: ", betaGenAdMitMH$accept, sprintf("\n"))
```

### Griddy Gibbs Sampler (GGS) 
```{r}
numGridPoints <- 25
lbParams <- apply(drawsRWMH, 2, function(x) {min(x) - 0.5 * stats::sd(x)})
lbParams[5] <- max(1E-5, lbParams[5]) # lbEtab dilimited to 0
ubParams <- apply(drawsRWMH, 2, function(x) {max(x) + 0.5 * stats::sd(x)})
ubParams[5] <- min(0.5 - 1E-5, ubParams[5]) # ubEtab dilimited to 0.5
ubParams[3] <- min(1 - 1E-5, ubParams[3])  # ubB to 1
grid <- apply(
  cbind(lbParams, ubParams),
  1,
  function(x) {seq(x[1], x[2], length.out = numGridPoints)}
)
startTime <- Sys.time()
drawsGGS <- GGS(
  "BetaGenTEGARCH",
  priorStack,
  y = as.matrix(returns),
  f1 = ltScale,
  initParams = initParams,
  grid = grid,
  iter = iter,
  logOffset = logOffset,
  printIter = 1000
)
endTime <- Sys.time()
timeGGS <- difftime(endTime, startTime, units = 'secs')
cat("GGS Time: ", timeGGS, sprintf(" seconds\n"))
```

### Hamiltonian Monte Carlo (HMC)
```{r}
lb = c(-Inf, -Inf, -1., -Inf, 0., 0.)
ub = c(Inf, Inf, 1., Inf, 0.5, Inf)
startTime <- Sys.time()
firstWarmUpHMC <- HMC(
  "BetaGenTEGARCH",
  priorStack,
  y = as.matrix(returns),
  f1 = ltScale,
  initParams = initParams,
  iter = 100,
  mass = diag(kNumParams),
  stepsize = 0.01,
  integrationTime = 0.2,
  lb = lb,
  ub = ub
)
secondWarmUpHMC <- HMC(
  "BetaGenTEGARCH",
  priorStack,
  y = as.matrix(returns),
  f1 = ltScale,
  initParams = initParams,
  iter = 100,
  mass = solve(cov(firstWarmUpHMC[-(1:50),])),
  stepsize = 0.1,
  integrationTime = 2.,
  lb = lb,
  ub = ub
)
drawsHMC <- HMC(
  "BetaGenTEGARCH",
  priorStack,
  y = as.matrix(returns),
  f1 = ltScale,
  initParams = initParams,
  iter = iter,
  mass = solve(cov(secondWarmUpHMC[-(1:20),])),
  stepsize = 0.5,
  integrationTime = 2.0,
  lb = lb,
  ub = ub
)
endTime <- Sys.time()
timeHMC <- difftime(endTime, startTime, units = 'secs')
cat("HMC Time: ", timeHMC, sprintf(" seconds\n"))

```

## Comparing MCMC Methods 

```{r}
mcmcNames <- c("RWMH", "Griddy Gibbs", "AdMit", "HMC")
N <- kNumParams * length(mcmcNames)
parNames <- list(
  expression(tilde(omega)),
  "A",
  "B",
  expression(mu),
  expression(bar(eta)),
  expression(upsilon)
)
# First draw is weird, should be ignored
betaGenAdMitMH$draws[1,] <- betaGenAdMitMH$draws[2,]
samples <- matrix(cbind(drawsRWMH, drawsGGS, betaGenAdMitMH$draws, drawsHMC),
                  iter, N)
colNames <- c()
for (mcmcName in mcmcNames) {
  for (parName in parNames) {
    colNames <- c(colNames, paste(mcmcName, parName, sep = " "))
  }
}
colnames(samples) <- colNames
burn <- 1000
```

### Convergence 
```{r}
mcmcSamples <- coda::mcmc(samples, start = burn, end = iter)
GC.diagnostic <- coda::geweke.diag(mcmcSamples, 0.1, 0.5)
zVals <- matrix(GC.diagnostic$z, ncol = kNumParams, byrow = TRUE)
zVals <- data.frame(zVals, row.names = mcmcNames)
colnames(zVals) <- c("OmegaTilde", "A", "B", "Mu", "EtaBar", "Upsilon")
knitr::kable(zVals, caption = "Geweke convergence diagnostics")
knitr::kable(abs(zVals) < 1.96, caption = "Geweke convergence diagnostics test @ 5% significance")
```

### Descriptic statistics of MCMC draws
```{r}
knitr::kable(summary(mcmcSamples)$statistics)
```

### Effective sample sizes (ESSs) & factors
```{r}
timesMcmc <- c(timeRWMH, timeGGS, timeAdMit, timeHMC)
ESSs <- coda::effectiveSize(mcmcSamples)
ESSs <- matrix(ESSs, ncol = kNumParams, byrow = TRUE)
ESSs <- data.frame(ESSs, row.names = mcmcNames)
colnames(ESSs) <- c("OmegaTilde", "A", "B", "Mu", "EtaBar", "Upsilon")
knitr::kable(ESSs, caption = "Effective sample sizes")
knitr::kable(ESSs / as.numeric(timesMcmc), caption = "Effective sample sizes per second")
```

### Trace plots 
```{r}
PlotTraces(
  samples,
  mcmcNames[-1],
  parNames = list("B", expression(bar(eta))),
  iters = (10000:10500),
  cols = c(2, 4)
)
```

### Histograms
```{r}
PlotHists(
  samples,
  mcmcNames[-1],
  parNames = list("B", expression(bar(eta)), expression(upsilon)),
  burn = burn
)
```

### Joint distribtuion plots 
```{r, fig.width = 6, fig.height = 6, fig.align = 'center'}
paramPairs <- list(
  list(expression(tilde(omega)), "B"),
  list("A", "B"),
  list(expression(omega), "B"),
  list(expression(bar(eta)), expression(upsilon))
)
colnames(drawsHMC) <- parNames
names(lbParams) <- parNames
names(ubParams) <- parNames
omegaDrawsHMC <- drawsHMC[,paste(parNames[1])] * (1 - drawsHMC[,"B"])
drawsHMC_ <- cbind(drawsHMC, "expression(omega)" = omegaDrawsHMC)
lbParams_ <- c(lbParams, "expression(omega)" = 0.0)
ubParams_ <- c(ubParams, "expression(omega)" = 0.85)
PlotJointDists(drawsHMC_, paramPairs, lbParams_, ubParams_, burn = burn)
```

### Auto Correlation Function (ACF) plots 
```{r}
PlotACFs(
  samples,
  mcmcNames[-1],
  parNames = list("B", expression(bar(eta))),
  burn = burn
)
```

## Posterior of volatility 

```{r}
volsGenDraws <- matrix(0, iter - burn, length(returns))
for (i in 1:(iter - burn)) {
  betaGen$SetParams(as.vector(drawsHMC[burn + i,]))
  volsGenDraws[i, ] <- betaGen$VolFilter(returns, ltScale)
  if ((i > 0) && ((i %% 1e4) == 0)) {
    cat(sprintf("iter %i\n", i));
  }
}
# Draws for which EtaBar is close to zero will result in under/overflow issues
# in the gamma function. This is corrected for by calculating vols for these
# draws using the limiting GED volatility formula.
zeroix <- which(
  apply(volsGenDraws, 1, function(x) any(x == 0)) |
  apply(volsGenDraws, 1, function(x) any(is.na(x)))
)
for (i in zeroix) {
  betaGen$SetParams(drawsHMC[burn + i,])
  scale <- betaGen$Filter(returns, ltScale)
  volsGenDraws[i, ] <- (
    exp(scale) * (betaGen$Upsilon ^ (1 / betaGen$Upsilon)) *
    sqrt(gamma(3 / betaGen$Upsilon) /  gamma(1 / betaGen$Upsilon))
  )
}
```

### Highest Posterior Density (HPD) volatility plots
```{r, fig.width = 10, fig.height = 6, fig.align = 'center'}
zoomDate <- "2016-09-12"
PlotHPDOverTime(
  volsGenDraws,
  returns,
  dates,
  startDate = "2016-04-21",
  endDate = "2017-04-21",
  ylab = "Returns",
  statStr = "volatility",
  obsStr = "S&P 500 absolute returns",
  statsML = volsGen,
  zoomDate = zoomDate,
  xy1 = list(x = c(17058, 17165), y = c(39.5, 102)),
  xy2 = list(x = c(17058, 17165), y = c(18, 55))
)
```

### Some posterior statistics of volatility
```{r, collapse = FALSE}
zoomDateIdx <- which(dates == as.Date(zoomDate))
zoomDateDraws <- volsGenDraws[, zoomDateIdx]
```
Return the day before `r zoomDate`: `r returns[zoomDateIdx - 1]`  
Mean skewness whole sample: `r mean(moments::skewness(volsGenDraws))`  
Mean kurtosis whole sample: `r mean(moments::kurtosis(volsGenDraws))`  
Mean skewness on `r zoomDate`: `r mean(moments::skewness(zoomDateDraws))`  
Mean kurtosis on `r zoomDate`: `r mean(moments::kurtosis(zoomDateDraws))`  


## Beta-t-EGARCH comparison

### ML estimation Beta-t-EGARCH
```{r ML_estimation}
initParamsBetaT  <- c(
  omega = ltScale,
  A = 0.05,
  B = 0.95,
  mu = mean(returns),
  NuBar = 0.2
)
betaT <- FitML(
  model = new(BetaTEGARCH),
  initParams = initParamsBetaT,
  y = returns,
  f1 = ltScale,
  method = 'BFGS',
  control = list(maxit = 1e5),
  hessian = TRUE,
  verbose = TRUE
)
betaT$SetParams(betaT$ParamsML)
volsT <- betaT$VolFilter(returns, ltScale)
```

### HMC Beta-t-EGARCH
```{r}
kNumParamsBetaT <- 5
priorStackBetaT <- new(
  PriorStack,
  rep("ImproperUniform", kNumParamsBetaT),
  list(
    c(-Inf, Inf),
    c(-Inf, Inf),
    c(-1, 1),
    c(-Inf, Inf),
    c(0, 0.5)
  )
)
lbBetaT = c(-Inf, -Inf, -1., -Inf, 0.)
ubBetaT = c(Inf, Inf, 1., Inf, 0.5)
startTime <- Sys.time()
firstWarmUpHMCBetaT <- HMC(
  "BetaTEGARCH",
  priorStackBetaT,
  y = as.matrix(returns),
  f1 = ltScale,
  initParams = initParamsBetaT,
  iter = 100,
  mass = diag(kNumParamsBetaT),
  stepsize = 0.01,
  integrationTime = 0.2,
  lb = lbBetaT,
  ub = ubBetaT
)
secondWarmUpHMCBetaT <- HMC(
  "BetaTEGARCH",
  priorStackBetaT,
  y = as.matrix(returns),
  f1 = ltScale,
  initParams = initParamsBetaT,
  iter = 100,
  mass = solve(cov(firstWarmUpHMCBetaT[-(1:50),])),
  stepsize = 0.1,
  integrationTime = 2.,
  lb = lbBetaT,
  ub = ubBetaT
)
drawsHMCBetaT <- HMC(
  "BetaTEGARCH",
  priorStackBetaT,
  y = as.matrix(returns),
  f1 = ltScale,
  initParams = initParamsBetaT,
  iter = iter,
  mass = solve(cov(secondWarmUpHMCBetaT[-(1:20),])),
  stepsize = 0.5,
  integrationTime = 2.0,
  lb = lbBetaT,
  ub = ubBetaT
)
endTime <- Sys.time()
timeHMCBetaT <- difftime(endTime, startTime, units = 'secs')
cat("HMC Time: ", timeHMCBetaT, sprintf(" seconds\n"))
```



### Posterior of volatility Beta-t-EGARCH
```{r}
volsTDraws <- matrix(0, iter - burn, length(returns))
for (i in 1:(iter - burn)) {
  betaT$SetParams(as.vector(drawsHMCBetaT[burn + i,]))
  volsTDraws[i, ] <- betaT$VolFilter(returns, ltScale)
  if ((i > 0) && ((i %% 1e4) == 0)) {
    cat(sprintf("iter %i\n", i));
  }
}
```

### HPD volatility plots Beta-t-EGARCH
```{r, fig.width = 10, fig.height = 6, fig.align = 'center'}
zoomDate <- "2016-09-12"
PlotHPDOverTime(
  volsTDraws,
  returns,
  dates,
  startDate = "2016-04-21",
  endDate = "2017-04-21",
  ylab = "Returns",
  statStr = "volatility",
  obsStr = "S&P 500 absolute returns",
  statsML = volsT,
  zoomDate = zoomDate,
  xy1 = list(x = c(17058, 17165), y = c(26, 102)),
  xy2 = list(x = c(17058, 17165), y = c(18, 55)),
  modeCol = rgb(0,0,1,1), 
  MLCol = "red", 
  fillCol = rgb(0,0,1,1/4),
  borderCol = rgb(0,0,1,1/2)
)
```

## Miscelaneuos

### Value at Risk (VaR) histogram
```{r, fig.width = 6, fig.height = 4, fig.align = 'centre'}
VaRBetaGenTEGARCH <- function(params, vols, muIdx, etaBIdx, upsIdx, alpha=0.99){
  if (is.matrix(params)) {
    mu <- params[, muIdx]
    etaB <- params[, etaBIdx]
    ups <- params[, upsIdx]
  }else{
    mu <- params[muIdx]
    etaB <- params[etaBIdx]
    ups <- params[upsIdx]
  }

  xt <- qbeta(2 * (1 - alpha), 1 / (etaB * ups), 1 / ups)
  Finvxt <- (((1 - xt) / (etaB * xt)) ^ (1 / ups))
  if (is.matrix(vols)) {
    numDraws <- dim(vols)[1]
    numObs <- dim(vols)[2]
    VaR <- matrix(0, numDraws, numObs)
    for (i in 1:numObs) {
      VaR[, i] <- mu + vols[, i] * Finvxt
    }
  }else{
    VaR <- mu + vols * Finvxt
  }
  return(VaR)
}
VaR99 <- VaRBetaGenTEGARCH(
  drawsHMC[-(1:burn),],
  volsGenDraws,
  "expression(mu)",
  "expression(bar(eta))",
  "expression(upsilon)",
  alpha = 0.99
)
VaR99ML <- VaRBetaGenTEGARCH(betaGen$ParamsML, volsGen, 4, 5, 6, alpha = 0.99)
par(mfrow = c(1, 1), mar = c(4.5, 4.5, 2, 2))
h <- hist(VaR99[, zoomDateIdx], breaks = 9:50 * 4, plot = FALSE)
plot(h, main = "", xlab = "")
title(xlab = expression(VaR[99]), cex.lab = 1.2)
abline(v = VaR99ML[zoomDateIdx], lty = 2, col = "blue")
abline(v = mean(VaR99[, zoomDateIdx]), lty = 2, col = "red")
leg <- c(
  paste("ML estimate VaR 99% ~ ", sprintf("%.1f", VaR99ML[zoomDateIdx])),
  paste("Posterior Mean VaR 99% ~ ", sprintf("%.1f", mean(VaR99[,zoomDateIdx])))
)
legend("topright", legend = leg, bty = "n", col = c("blue", "red"), 
       lty = c(2, 2))
```

### Degrees of freedom parameter histogram
```{r, fig.width = 6, fig.height = 4, fig.align = 'centre'}
drawsEta <- 1 / drawsHMC[-(1:burn), "expression(bar(eta))"]
cat("Posterior range d.o.f. parameter Beta-Gen-t-EGARCH: ", range(drawsEta), 
    sprintf("\n"))
cat("Posterior range d.o.f. parameter Beta-t-EGARCH: ", 
    range(1 / drawsHMCBetaT[,5]), sprintf("\n"))
hist(main = "", drawsEta[drawsEta < 200], xlab = "", breaks = 0:40 * 5)
title(xlab = expression(eta), cex.lab = 1.2)
```

### Score function plots
```{r, fig.width = 5, fig.height = 5, fig.align = 'centre'}
scoreGenT <- function(etaBar, upsilon, residual){
  b <- (etaBar * (abs(residual) ^ upsilon) /
          (etaBar * (abs(residual) ^ upsilon) + 1))
  return(((etaBar + 1) / etaBar) * b - 1)
}
scoreGED <- function(upsilon, residual){
  return((abs(residual) ^ upsilon) - 1)
}
drawsA <- drawsHMC[-(1:burn), "A"]
drawsMu <- drawsHMC[-(1:burn), "expression(mu)"]
drawsEtaB <- drawsHMC[-(1:burn), "expression(bar(eta))"]
drawsUps <- drawsHMC[-(1:burn), "expression(upsilon)"]
par(mfrow = c(1, 1), mar = c(4.5, 4.5, 2, 2))
ixMin <- which.min(volsGenDraws[, zoomDateIdx])
ixMax <- which.max(volsGenDraws[, zoomDateIdx])
fGenDraws <- matrix(0, iter - burn, length(returns))
for (i in 1:(iter - burn)) {
  betaGen$SetParams(as.vector(drawsHMC[burn + i,]))
  fGenDraws[i, ] <- betaGen$Filter(returns, ltScale)
  if ((i > 0) && ((i %% 1e4) == 0)) {
    cat(sprintf("iter %i\n", i));
  }
}
epsMin <- (exp(-fGenDraws[, zoomDateIdx - 1][ixMin]) *
             (returns[zoomDateIdx - 1] - drawsMu[ixMin]))
epsMax <- (exp(-fGenDraws[, zoomDateIdx - 1][ixMax]) *
             (returns[zoomDateIdx - 1] - drawsMu[ixMax]))
xVals <- -25:25 / 2.5
scoreGenMin <- scoreGenT(drawsEtaB[ixMin], drawsUps[ixMin], xVals)
scoreGenMax <- scoreGenT(drawsEtaB[ixMax], drawsUps[ixMax], xVals)
scoreGenMaxAdj <- (scoreGenT(drawsEtaB[ixMax], drawsUps[ixMax], xVals) *
                     drawsA[ixMax] / drawsA[ixMin])
scoreGenEpsMin <- scoreGenT(drawsEtaB[ixMin], drawsUps[ixMin], epsMin)
scoreGenEpsMax <- scoreGenT(drawsEtaB[ixMax], drawsUps[ixMax], epsMax)
scoreGenEpsMaxAdj <- (scoreGenT(drawsEtaB[ixMax], drawsUps[ixMax], epsMax) *
                     drawsA[ixMax] / drawsA[ixMin])
plot(
  xVals,
  scoreGenMin,
  ylab = "Score",
  xlab = "Standardized observation",
  col = "red",
  type = "l",
  ylim = c(-1.5, 25)
)
lines(xVals, scoreGenMax, col = "blue")
lines(xVals, scoreGenMaxAdj, col = "blue", lty = 2)
points(epsMin, scoreGenEpsMin, col = "green", pch = 8)
points(epsMax, scoreGenEpsMax, col = "magenta", pch = 8)
points(epsMax, scoreGenEpsMaxAdj, col = "magenta", pch = 8)
leg <-
  c(
    "Score function min",
    "Score function max",
    "Score function max adj. for A",
    "Standardized obs. 2016-09-09 min",
    "Standardized obs. 2016-09-09 max"
  )
legend(
  "topright",
  legend = leg,
  col = c("red", "blue", "blue", "green", "magenta"),
  lty = c(1, 1, 2, NA, NA),
  pch = c(NA, NA, NA, 8, 8)
)
par(mfrow = c(1, 1), mar = c(4.5, 2, 2, 2))
plot(xVals, scoreGenT(0.2, 2, xVals), col = "red", type = "l",
     ylim = c(-1.5, 10), xlab = "Standardized observation")
lines(xVals, scoreGED(1, xVals), col = "blue")
lines(xVals, scoreGED(1.5, xVals), col = "green")
leg <- c("Student-t, 5 d.o.f.", "Laplace", "Gaussian")
legend("topright", legend = leg, col = c("red", "blue", "green"), 
       lty = c(1, 1, 1))

```

